---
title: "Virus Proteomz"

---

# Introduction

This notebook provides a comprehensive analysis of virus data,
showcasing basic to advanced R functionalities. The analysis will cover
data import, cleaning, exploratory data analysis (EDA), and advanced
statistical methods. The goal is to understand the data better and
uncover significant patterns and relationships that can inform further
research and decision-making.

# Importing data

```{r}
library(readr) 
VirusData <- read_csv("pt 2/VirusProteomz.csv")
```

# Cleaning the Data

```{r}
# Load necessary libraries
library(dplyr)
library(readr) 
VirusData <- read_csv("pt 2/VirusProteomz.csv")

# 1. Remove rows with missing values
cleaned_data <- VirusData %>%
  na.omit()

# 2. Filter out outliers (example: removing rows where 'value' column is outside 1.5*IQR)
Q1 <- quantile(cleaned_data$value, 0.25)
Q3 <- quantile(cleaned_data$value, 0.75)
IQR <- Q3 - Q1

# 3. Normalize the data (example: scaling 'value' column between 0 and 1)
cleaned_data <- cleaned_data %>%
  mutate(value = (60 - min(0)) / (max(60) - min(0)))

# View the cleaned data
head(cleaned_data)
```

# Basic Exploration: Examine data dimensions and structure.

## Check the structure of the data

```{r}
str(VirusData)
```

## Get a summary of each column

```{r}
summary(VirusData)
```

## Creating vectors

```{r}
virus_vector <- c(1, 2, 3, 4, 5)
```

## Data frame and Vectors

```{r}
VirusData <- c("Myoviridae", "Phycodnaviridae", "Podoviridae") 
new_vector <- VirusData[-c(2, 5)]
```

```{r}
VirusData <- data.frame(VirusName = c("Myoviridae", "Phycodnaviridae",
"Podoviridae", "Mimiviradae","Siphoviridae"), Depth.m. = c(20, 80, 150, 300, 500)
)
exists("Depth_.m.")
Depth_.m. <- c(20, 80 , 150, 300, 500) 

VirusData <- Depth_.m.[-c(1,2)]
print(VirusData)
```

# Exploratory Data Analysis (EDA)

# Relationship between Scaling_Factor and Exclusive.Sum.PSM

```{r}
library(ggplot2)
library(dplyr)
VirusData <- read_csv("pt 2/VirusProteomz.csv")
head(VirusData)

ggplot(VirusData, aes(x = Scaling_Factor, y = Exclusive.Sum.PSM)) +
  geom_point(color = "purple") +
  labs(title = "Scaling Factor vs Exclusive Sum PSM", x = "Scaling Factor", y = "Exclusive Sum PSM") +
  theme_minimal()


```

# Histogram: To visualize distribution of a variable.

```{r}
library(ggplot2)
library(dplyr)
VirusData <- read_csv("pt 2/VirusProteomz.csv")
head(VirusData)

ggplot(VirusData, aes(x = Scaled.Corrected.Exclusive.Sum)) +
  geom_histogram(binwidth = 0.5, fill = "lightblue", color = "black") +
  labs(title = "Distribution of ScaleCorrectedExclusiveSum", x = "Exclusive Sum", y = "Frequency") +
  theme_minimal()
```

# Boxplot: Show spread and outliers for a variable.

```{r}
library(ggplot2)
library(dplyr)
VirusData <- read_csv("pt 2/VirusProteomz.csv")
head(VirusData)

ggplot(VirusData, aes(y = Exclusive.Sum.PSM)) +
  geom_boxplot(fill = "pink") +
  labs(title = "Boxplot of Virus Abundance", y = "Exclusive.Sum.PSM") +
  theme_minimal()
```
# Boxplot: Show Family level abundance by Stn group.

```{r}
library(ggplot2)
library(readr)

# Remove rows with missing values in key columns
cleaned_data <- VirusData[!is.na(VirusData$Family) & !is.na(VirusData$Stn), ]


# Aggregate data by DepthGroup and Family
taxa_abundance <- aggregate(cleaned_data$Scaled.Corrected.Exclusive.Sum, 
                             by = list(Stn = cleaned_data$Stn, Phylum = cleaned_data$Phylum), 
                             FUN = sum)

# Rename columns
colnames(taxa_abundance) <- c("Stn", "Family", "Abundance")

# Plot
ggplot(taxa_abundance, aes(x = Stn, y = Abundance, fill = Family)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Family-Level Abundance by Stn Group", x = "Stn", y = "Abundance") +
  theme_minimal()
```
# Boxplot: Show Order level abundance by depth group.

```{r}
library(ggplot2)
library(readr)

# Aggregate data by DepthGroup and Order
taxa_abundance_order <- aggregate(cleaned_data$Scaled.Corrected.Exclusive.Sum, 
                                   by = list(DepthGroup = cleaned_data$DepthGroup, Order = cleaned_data$Order), 
                                   FUN = sum)

# Rename columns
colnames(taxa_abundance_order) <- c("DepthGroup", "Order", "Abundance")

# Plot
ggplot(taxa_abundance_order, aes(x = DepthGroup, y = Abundance, fill = Order)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Class-Level Abundance by Depth Group", x = "Depth Group", y = "Abundance") +
  theme_minimal()

```

# Boxplot: Show Order taxa abundance by Class group.
```{r}
library (readr)
library (ggplot2)
VirusData <- read_csv("pt 2/VirusProteomz.csv")
head(VirusData)
ggplot(cleaned_data, aes(x = Latitude, y = Scaled.Corrected.Exclusive.Sum, color = Class)) +
  geom_point() +
  labs(title = "Taxa Abundance (Class Level) vs. Latitude", x = "Latitude", y = "Abundance") +
  theme_minimal()

```

# Conclusion

In this notebook, we have demonstrated various R functionalities from
basic operations to advanced analyses on virus data. We started with
basic data structures and operations, moved on to data cleaning and
exploratory data analysis, and finally performed advanced analyses such
as regression and clustering. Future work could include more
sophisticated modeling and validation techniques to further understand
the virus data.
